Concurrent Programming in Harmony

### 2021-02-23 book/paper.tex
Latest

@@

\begin{document}
\maketitle

@@

Harmony can report the following failure types:
\begin{itemize}
\item \texttt{Safety violation}: this means something went wrong with
at least one of the executions of the program that it tried.  This
can include a failing assertion, divide by zero, using an uninitialized
or non-existent variable, dividing a set by an integer, and so on.
% It also includes certain infinite loops.
Harmony will print a trace of the shortest bad execution that it found.
\item \texttt{Non-terminating States}: Harmony found one or more states
from which there does not exist an execution such that all threads
terminate.  Harmony will not only print the non-terminating state with
the shortest trace, but also the list of threads
at that state, along with their program counters.
% \item \texttt{Stopped States}: Similar to non-terminating states,
% these are states in which some threads are left that cannot
% terminate.  \autoref{ch:synch} will explain what it means for
% a thread to be stopped.
\end{itemize}

@@

In 1981, Gary L.~Peterson came up with a beautiful solution to the mutual exclusion
problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
The algorithm is an amalgam of the (incorrect) algorithms in
\autoref{fig:upflags} and \autoref{fig:upturn}, and is presented
in \autoref{fig:peterson}.
A thread first indicates its interest in entering the critical section
by setting its flag.
It then politely gives way to the other thread should it also want to
enter the critical section---if both do so at the same time one will
win because writes to memory in Harmony are atomic.
The thread continues to be polite, waiting
until either the other thread is nowhere near the critical section
($\mathit{flag}[1 - \mathit{self}] == \mathtt{False}$) or has given way
($\mathit{turn} == \mathit{self}$).
Running the algorithm with Harmony shows that it satisfies both mutual
exclusion and progress.

@@

\section*{Exercises}
\begin{problems}
\item \autoref{fig:csonebit} presents another correct solution to the
mutual exclusion problem.  It is similar to the one in
\autoref{fig:upflags}, but has a thread \emph{back out and try again}
if it finds that the other thread is either trying to enter the critical
section or already has.  Compare this algorithm with Peterson's.  What are
the advantages and disadvantages?
\item
Can you find one or more inductive invariants for the algorithm in
\autoref{fig:csonebit} to prove it correct?
Here's a pseudo-code version of the algorithm to help you.  Each line
is an atomic action:
\begin{code}
\begin{verbatim}
    initially: flagX = flagY = False

@@

{\small
\begin{tabular}{|l|l|}
\hline
Address & compute address from two components \\
\hline
Apply & pop $m$ and $i$ and apply $i$ to $m$, pushing a value \\
\hline
Assert, Assert2 & pop $b$ and check that it is \texttt{True}.  Assert2 also pops value to print \\
\hline
AtomicInc/Dec & increment/decrement the atomic counter of this context \\
\hline
Continue & no-op (but causes a context switch) \\
\hline
Choose & choose an element from the set on top of the stack \\
\hline
Cut & cut a set into its smallest element and the remainder \\
\hline
Del [$v$] & delete shared variable $v$ \\
\hline
DelVar [$v$] & delete thread variable $v$ \\
\hline
Dup & duplicate the top element of the stack \\
\hline
Frame \texttt{m} $a$ & start method \texttt{m} with arguments $a$,
initializing variables.  \\
\hline
Go & pop context and value, push value on context's stack, and add to context bag \\
\hline
IncVar $v$ & increment thread variable $v$ \\
\hline
Jump $p$ & set program counter to $p$ \\
\hline
JumpCond $e$ $p$ & pop expression and, if equal to $e$, set program counter to $p$ \\
\hline
Load [$v$] & push the value of a shared variable onto the stack \\
\hline
LoadVar [$v$] & push the value of a thread variable onto the stack \\
\hline
Move $i$ & move stack element at offset $i$ to top of the stack \\
\hline
$n$-ary \textit{op} & apply $n$-ary operator \textit{op} to the top $n$ elements on the stack \\
\hline
Pop & pop a value of the stack and discard it \\
\hline
Push $c$ & push constant $c$ onto the stack \\
\hline
ReadonlyInc/Dec & increment/decrement the read-only counter of this context \\
\hline
Return & pop return address, push \texttt{result}, and restore program counter \\
\hline
SetIntLevel & pop $e$, set interrupt level to $e$, and push old interrupt level \\
\hline
Spawn & pop initial thread-local state, argument, and method and spawn a new context \\
\hline
Split & pop tuple and push its elements \\
\hline
Stop [$v$] & save context into shared variable $v$ and remove from context bag \\
\hline
Store [$v$] & pop a value from the stack and store it in a shared variable \\
\hline
StoreVar [$v$] & pop a value from the stack and store it in a thread variable \\
\hline
Trap & pop interrupt argument and method \\
\hline
\end{tabular}
}

@@

In Python, lists are objects.  Thus $x$ and $y$ point to the same list,
and the assertion would fail if executed by Python.
In Harmony, lists are values.  So when $x$ is updated in Line~3, it does
not affect the value of $y$.  The assertion succeeds.
Harmony supports references to values (\autoref{ch:method}),
allowing programs to implement shared objects.

>>>>

\newglossaryentry{sequential consistency}
  name=sequential consistency,
  description={is a consistency model in which shared memory accesses are
    executed in an order consistent with the program order}

@@

\begin{document}
\begin{titlepage}
    \centering
    \vfill
    {\bfseries\huge
        Concurrent Programming in Harmony
    }
    \vfill
    \includegraphics[width=6cm]{figures/harmony-icon.png}
    \vfill
    \vfill
    \hfill \Large Robbert van Renesse
\end{titlepage}

@@

Harmony can report the following failure types:
\begin{itemize}
\item \texttt{Safety violation}: This means something went wrong with
at least one of the executions of the program that it tried.  This
can include a failing assertion, divide by zero, using an uninitialized
or non-existent variable, dividing a set by an integer, and so on.
% It also includes certain infinite loops.
Harmony will print a trace of the shortest bad execution that it found.
\item \texttt{Non-terminating State}: Harmony found one or more states
from which there does not exist an execution such that all threads
terminate.  Harmony will not only print the non-terminating state with
the shortest trace, but also the list of threads
at that state, along with their program counters.
\item \texttt{Active Busy Waiting}:  There are states in which
some thread cannot make progress without the help of another thread but
does not block;
\item \texttt{Data Race}: There are states in which two or more threads
concurrently access a shared variable, at least one of which is a store
operation.
% \item \texttt{Stopped States}: Similar to non-terminating states,
% these are states in which some threads are left that cannot
% terminate.  \autoref{ch:synch} will explain what it means for
% a thread to be stopped.
\end{itemize}

@@

\glsadd{sequential consistency}%
In 1981, Gary L.~Peterson came up with a beautiful solution to the mutual exclusion
problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
The algorithm is an amalgam of the (incorrect) algorithms in
\autoref{fig:upflags} and \autoref{fig:upturn}, and is presented
in \autoref{fig:peterson}.
(The first line specifies that the \textit{flags} and \textit{turn}
variables should have \emph{sequential consistency}---it prevents
Harmony from complaining about data races involving these variables,
explained in \autoref{ch:synch}.)

A thread first indicates its interest in entering the critical section
by setting its flag.
It then politely gives way to the other thread should it also want to
enter the critical section---if both do so at the same time one will
win because writes to memory in Harmony are atomic.
The thread continues to be polite, waiting
until either the other thread is nowhere near the critical section
($\mathit{flag}[1 - \mathit{self}] == \mathtt{False}$) or has given way
($\mathit{turn} == \mathit{self}$).
Running the algorithm with Harmony shows that it satisfies both mutual
exclusion and progress.

@@

\section*{Exercises}
\begin{problems}
\item \autoref{fig:csonebit} presents another solution to the
mutual exclusion problem.  It is similar to the one in
\autoref{fig:upflags}, but has a thread \emph{back out and try again}
if it finds that the other thread is either trying to enter the critical
section or already has.  Compare this algorithm with Peterson's.
Why does Harmony complain about \emph{active busy waiting}?
\item
Can you find one or more inductive invariants for the algorithm in
\autoref{fig:csonebit} to prove it correct?
Here's a pseudo-code version of the algorithm to help you.  Each line
is an atomic action:
\begin{code}
\begin{verbatim}
    initially: flagX = flagY = False

@@

{\small
\begin{tabular}{|l|l|}
\hline
Address & compute address from two components \\
\hline
Apply & pop $m$ and $i$ and apply $i$ to $m$, pushing a value \\
\hline
Assert, Assert2 & pop $b$ and check that it is \texttt{True}.  Assert2 also pops value to print \\
\hline
AtomicInc/Dec & increment/decrement the atomic counter of this context \\
\hline
Continue & no-op (but causes a context switch) \\
\hline
Choose & choose an element from the set on top of the stack \\
\hline
Cut & cut a set into its smallest element and the remainder \\
\hline
Del [$v$] & delete shared variable $v$ \\
\hline
DelVar [$v$] & delete thread variable $v$ \\
\hline
Dup & duplicate the top element of the stack \\
\hline
Frame \texttt{m} $a$ & start method \texttt{m} with arguments $a$,
initializing variables.  \\
\hline
Go & pop context and value, push value on context's stack, and add to context bag \\
\hline
IncVar $v$ & increment thread variable $v$ \\
\hline
Jump $p$ & set program counter to $p$ \\
\hline
JumpCond $e$ $p$ & pop expression and, if equal to $e$, set program counter to $p$ \\
\hline
Load [$v$] & push the value of a shared variable onto the stack \\
\hline
LoadVar [$v$] & push the value of a thread variable onto the stack \\
\hline
Move $i$ & move stack element at offset $i$ to top of the stack \\
\hline
$n$-ary \textit{op} & apply $n$-ary operator \textit{op} to the top $n$ elements on the stack \\
\hline
Pop & pop a value of the stack and discard it \\
\hline
Push $c$ & push constant $c$ onto the stack \\
\hline
ReadonlyInc/Dec & increment/decrement the read-only counter of this context \\
\hline
Return & pop return address, push \texttt{result}, and restore program counter \\
\hline
Sequential & pop an address of a variable that has sequential consistency \\
\hline
SetIntLevel & pop $e$, set interrupt level to $e$, and push old interrupt level \\
\hline
Spawn & pop initial thread-local state, argument, and method and spawn a new context \\
\hline
Split & pop tuple and push its elements \\
\hline
Stop [$v$] & save context into shared variable $v$ and remove from context bag \\
\hline
Store [$v$] & pop a value from the stack and store it in a shared variable \\
\hline
StoreVar [$v$] & pop a value from the stack and store it in a thread variable \\
\hline
Trap & pop interrupt argument and method \\
\hline
\end{tabular}
}

@@

In Python, lists are objects.  Thus $x$ and $y$ point to the same list,
and the assertion would fail if executed by Python.
In Harmony, lists are values.  So when $x$ is updated in Line~2, it does
not affect the value of $y$.  The assertion succeeds.
Harmony supports references to values (\autoref{ch:method}),
allowing programs to implement shared objects.
